Synthesizing Symmetric Lenses

### 2019-01-29 papers/SymmetricLenses/main.tex
Polishing on intro and beginning of 2

\textit{Synthesizing} lenses 
can make the task even easier.  Indeed, lens languages are excellent
targets for search-based synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
limit the space of possible programs. 
%
In particular, the Optician~\cite{optician} system has demonstrated that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, Optician's synthesis algorithm is guaranteed
to find a bijective lens matching the examples, if one exists. Indeed, 
it can often find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

@@

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly the symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  To check that they are expressive
enough in practice, we add simple symmetric lenses to the Boomerang
language~\cite{Boomerang} and apply them to a range of real-world
applications, showing in the process that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

However, we are still left with a second difficulty with synthesizing
symmetric lenses: While the number of bijective lenses between two related
formats is typically tiny, the number of simple symmetric lenses is
typically huge.  Thus, if a na\"ive search algorithm selects the first
simple symmetric lens it finds, this will generally \emph{not} be the one we
want.  We need a new principle for identifying ``more likely'' lenses and a
more sophisticated synthesis algorithm that uses this principles to search
the space more intelligently.

For these, we turn to information theory. 
We consider a ``likely'' lens to be one that propagates ``a lot'' of
information from source to 
target and vice versa. Conversely, an unlikely lens requires a
large amount of additional information to recover the target given a
source.  
This categorization follows from the intuition
that users will typically want to synchronise formats with lots
of shared information, and so lenses that propagate
more information are preferable to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the shared information content of
two formats\bcp{It's not obvious at this point what the {\em shared}
  information content has to do with anything.} and our measure of likelihood.

@@

With simple symmetric lenses and our SRE-based likelihood measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likelihood measure.
The interesting complications arise from the need to deal with
regular-expression equivalences.  There are infinitely many regular
expressions equivalent to a given one, and the lens returned by a
type-directed search will depend on which of the infinitely many possible
representations are chosen for its source and target formats.  Moreover,
there may be {\em no} lens connecting the 
source and target formats selected by the user: in general, the synthesis
algorithm has to search to find equivalent ones for which a suitable lens
exists.  
To tame this complexity, we divide the synthesis algorithm into two
communicating search procedures (Figure~\ref{fig:high-level-algorithm}),
following~\citet{optician}. 
The first, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second, \GreedySynth, takes a pair of SREs as input
and uses a greedy, type-directed algorithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The whole synthesis algorithm terminates when comparing the
likilihood score of the best lens found so far with 
the complexity of newly generated SREs 
suggests that further searching is unlikely to be fruitful.

@@

We added this synthesis procedure to the Boomerang implementation and evaluated
its effectiveness on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).
The ``programmer'' can control the search process via examples and annotations
that informs the synthesis algorithm as to whether certain bits of information
are irrelevant, or must maintained.

In summary, our contributions are:\bcp{The section references do not proceed
  in order, so this doesn't work very well as an outline of the paper.
  Should we try to reorder the contributions?  Write a separate outline?
  Live with it as-is?}\afm{Personally, I don't see it as a problem.  It's
  nice when the contributions mirror the paper, I am not jarred when they
  don't.}\bcp{But then the reader has no birds-eye view of the paper at
  this point.  I tried reorganizing, which I think also helps clarify the
  contributions.} 
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We
present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give
an algorithm for synthesizing lenses built from these combinators, using a 
novel application of stochastic regular expressions to guide the search
process (\S\ref{sec:synthesis})\bcp{``and prove it correct''???}.  
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when applied to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices\bcp{The appendices are pretty
  hideous and hard to navigate!!}
\ifappendices\else included with the supplemental material\bcp{Right?}\fi.

@@

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\vskip 1.5ex
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \tag{1}
 \mbox{$\CreateR \OfType X \rightarrow Y$}\\
 \tag{2}
 \mbox{$\CreateL \OfType Y \rightarrow X$}\\
 \tag{3}
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$}\\
 \tag{4}
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$}
\end{gather} 
\end{minipage} 
\begin{minipage}[c]{.6\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\vskip 1ex
\bcp{I want less space between the put/create laws and their names.}
\noindent
The two \KW{create} functions are used to propagate shared information and
fill in default values when introducing new data (\EG, the ``unk'' salary
entry is added alongside the name to the management file when HR inserts a
new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the
other.  The round-tripping laws guarantee that pushing an ``un-edited''
value from one side (the result of a put or create) back through a lens in
the other direction will get back to exactly where we began.

@@

Generally, we do not expect programmers to define simple symmetric lenses by
writing four functions by hand and showing they satisfy the round-tripping
laws. Instead, we provide a set of combinators---a domain-specific
language---for building complex lenses from simpler ones.  Such languages
have been given for building lenses over many different kinds of data,
including database relations~\cite{BohannonPierceVaughan} and algebraic data
types~\cite{symmetric-lenses}\bcp{and what else? graphs, etc.?}.  We focus
here on combinators for defining lenses between string data formats
described by regular expressions.

>>>>

\textit{Synthesizing} lenses 
can make the task even easier.  Indeed, lens languages are excellent
targets for search-based synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
limit the space of possible programs. 
%
In particular, the Optician~\cite{optician} algorithm\bcp{there's a
  potential for confusion between ``Optician'' and ``Boomerang''---are they
  distinct systems?  Is one part of the other?  Is one a system and one just
  an algorithm?  It's not very clear.} has demonstrated that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, Optician's synthesis algorithm is guaranteed
to find a bijective lens matching the examples, if one exists. Indeed, 
it can often find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

@@

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly the symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  To check that they are expressive
enough in practice, we add simple symmetric lenses to the Boomerang
language~\cite{Boomerang}\bcp{this seems to assume that readers already know what
  Boomerang is} and apply them to a range of real-world
applications, showing in the process that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

This leaves us with just the second difficulty with synthesizing symmetric
lenses: While the number of bijective lenses between two given formats is
typically tiny, the number of simple symmetric lenses is typically enormous.
If a na\"ive search algorithm just selects the first simple symmetric lens
it finds, this will generally \emph{not} be the one the user wanted.  We
need a new principle for identifying ``more likely'' lenses and a more
sophisticated synthesis algorithm that uses this principle to search the
space more intelligently. For these, we turn to information theory.

We consider a ``likely'' lens to be one that propagates ``a lot'' of
information from source to 
target and vice versa. Conversely, an unlikely lens is one that requires a
large amount of additional information to recover the target given a
source.\bcp{I'm nervous about the way we're talking about source and target
  here.  For one thing, even the words ``source'' and ``target'' connote a
  directionality that is misleading here.  And even ignoring that issue, a
  lot of the discussion (like here) sounds as though we are talking about
  unidirectional transformations.}  
This categorization follows from the intuition
that users will typically want to synchronise formats with lots
of shared information\bcp{really?  It's hard to be certain without lots of
  practical experience, but a priori it seems equally reasonable for users
  to want to pull little bits of shared information out of huge structures
  and keep them in sync.  But maybe we can weaken the claim and just say
  that we have an intuition that this will at least be {\em one} large class
  of application scenarios, and this is the one we focus on}, and so lenses
that propagate more information are preferable to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the shared information content of
two formats and our measure of likelihood.

@@

With simple symmetric lenses and our SRE-based likelihood measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likelihood measure.
The interesting complications arise from the need to deal with
regular-expression equivalences.  There are infinitely many regular
expressions equivalent to a given one, and the lens returned by a
type-directed search will in general depend on which of the possible
representations are chosen for its source and target formats.  Moreover,
there may be {\em no} lens connecting the 
source and target formats selected\bcp{until we get past the colon, it is
  not clear that we are talking here about formats {\em not} modulo
  equivalence.} by the user: in general, the synthesis 
algorithm has to search to find equivalent ones for which a suitable lens
exists.  
To tame this complexity, we divide the synthesis algorithm into two
communicating search procedures (Figure~\ref{fig:high-level-algorithm}),
following~\citet{optician}. 
The first, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second, \GreedySynth, takes a pair of SREs as input
and uses a greedy, type-directed algorithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The whole synthesis algorithm terminates when a comparison of the
likilihood score of the best lens found so far with 
the complexity\bcp{what does it mean to compare likelihood with complexity?}
of newly generated SREs  
suggests that further searching is unlikely to be fruitful.

@@

We have added this synthesis procedure to the Boomerang system and evaluated
its effectiveness on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).
The ``programmer''\bcp{user?} can control the search process via examples and annotations
that informs the synthesis algorithm as to whether certain bits of information
are irrelevant, or must maintained.

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We
present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give
an algorithm for synthesizing lenses built from these combinators, using a 
novel application of stochastic regular expressions to guide the search
process (\S\ref{sec:synthesis})\bcp{``and prove it correct''!  (right?)}.  
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices\bcp{The appendices are pretty
  hideous and hard to navigate!!}
\ifappendices\else included with the supplemental material\bcp{Right?}\fi.

@@

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\vskip 1.5ex
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \tag{1}
 \mbox{$\CreateR \OfType X \rightarrow Y$}\\
 \tag{2}
 \mbox{$\CreateL \OfType Y \rightarrow X$}\\
 \tag{3}
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$}\\
 \tag{4}
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$}
\end{gather} 
\end{minipage} 
\begin{minipage}[c]{.6\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\vskip 1ex
\bcp{I want less space between the put/create laws and their names.}
\noindent
The two \KW{create} functions are used to propagate shared information and
fill in default values when introducing new data (\EG, the ``unk'' salary
entry is added\bcp{is added by what?  A create function, I guess, but the
  connection is not clear} alongside the name to the management file when HR inserts a
new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the
other.  The round-tripping laws guarantee that pushing an ``un-edited''
value from one side (the result of a put or create) back through a lens in
the other direction will get back to exactly where we began.  \bcp{That
  paragraph seems likely to be quite opaque to any reader that is not
  already familiar with (symmetric) lenses.  How about a concrete example?}

@@

We do not expect programmers to define simple symmetric lenses by
writing four functions by hand and showing they satisfy the round-tripping
laws.  Instead, we provide a set of combinators---a domain-specific
language---for building complex lenses from simpler ones.  Such languages
have been given for building lenses over many different kinds of data,
including database relations~\cite{BohannonPierceVaughan} and algebraic data
types~\cite{symmetric-lenses}\bcp{and what else? graphs, etc.?}.  We focus
here on combinators for defining lenses between string data formats
described by regular expressions.
